### 2019-10-01 ch1_introduction/content.tex
clean up thesis
update text

\section{The cell}
The cell is the smallest unit of life, of which all known living organisms are composed. Every cell houses a plethora of biomolecular processes that allow it to continuously adapt to changes in its environment. Due to the dynamic nature of these processes, it can be very challenging to comprehend the cellular response to a signal. A reductionist approach to understanding a complex biological system is to study the biochemical components of which it is comprised\cite{brigandt_reductionismbiology_2017}.

Recent advances in experimental technologies are playing a crucial role in reductionist biology, allowing to measure the abundance of thousands of different biochemical molecules in tens of thousands of individual cells. With it comes the challenge of analysing a large amounts of data that are not easily interpretable by hand. The sheer volume of the data generated from such highly-integrative and high-throughput experiments are not the only reason why they are so challenging to interpret. For instance, the generated data contains high levels of noise arising from inherent biomolecular stochasticity in the cells and from the experimental profiling techniques used, as well as batch effects arising from differences between donors and labs \cite{hon_humancellatlas_2018}. Biologists thus turn to computer scientists
to develop new tools to tackle these problems and help them to extract meaningful biological insights from the data. In this work, incremental contributions were made to the field in order to be able to address the aforementioned problems in a more comprehensive context.

Observing the biomolecular insides of cells can ultimately provide fundamental insights into the processes that govern these cells and help uncover novel approaches for disease diagnosis, prognosis, and treatment. For example, the Human Cell Atlas (HCA) consortium \cite{regev_humancellatlas_2018} has set out to develop a comprehensive reference map of all the different types of cells in the human body. Experts in the field often metaphorically describe the HCA initiative as aiming to develop a 'Google Maps' of the human body. Even in its infancy, the HCA has profiled 3.8 million cells from 248 donors across 42 labs \cite{humancellatlasconsortium_humancellatlas_2018}, and this number is likely to increase well above one hundred million.

@@

\subsection{The origins of life and the RNA world}
The discovery of the double helix shape of deoxyribonucleic acid (DNA) \cite{watson_molecularstructurenucleic_1953} is often considered the pivot point in our understanding of the origins of life and evolution. By now, it is well known that DNA serves as a medium for storing the genetic information required to reproduce a whole organism. With other words, the DNA of an organism contains the complete set of instructions required to build all of the biomolecular machinery present in its body.

Even so, a widely-accepted hypothesis states that life (or cells) did not originate from DNA, but instead was kicked off from its lesser-known cousin, ribonucleic acid (RNA). According to the RNA world hypothesis \cite{alberts_rnaworldorigins_2002}, the very first primitive cells used RNA both to store genetic information and perform the chemical reactions required to sustain themselves (Figure \ref{fig:rnaworld}). Only later did cells develop the ability to use the more chemically stable DNA molecules to self-sustain in a process commonly referred to as the Central Dogma.

\begin{figure}[htb!]
	\centering\includegraphics[width=.5\linewidth]{fig/rnaworld} % TODO: change image
	\caption{
		\textbf{RNA world.} The postulated rise and fall of the RNA world during the evolution of life, from early self-replicating RNA to complex, RNA-controlled metabolism, to the invention of translation, followed by diversification of all modern branches of life. Image from Horning (2011) \cite{horning_rnaworld_2011}. % TODO: expand the description
	}
	\label{fig:rnaworld}
\end{figure}

The central dogma describes the general flow of genetic information in almost all existent living cells: DNA is decoded to RNA, which in turn encodes proteins. Main processes involved in the central dogma are transcription, splicing and translation (Figure \ref{fig:centraldogma}).

@@

A stretch of DNA that contains a genetic blueprint for a particular molecule is called a gene, and the collection of an organism's genes is called its genome. \textbf{Transcription} is the process of synthesising an RNA molecule from a gene, and the resulting molecule is called a transcript. RNA is similar to DNA but differs in several ways; most notably all thymine nucleobases are replaced with uracil (U), and RNA molecules consist of only one strand.
% HT: Add: which is able to exit the nucleus compartment.
% RC: I haven't introduced the concept of a nucleus yet
Due to its single-strandedness, RNA is less stable and will break down faster. Single-strandedness also allows some types of RNA (e.g. transfer RNA, ribosomal RNA) to form more complex three-dimensional structures by having certain regions bind to other regions of the strand. This work only considers messenger RNA (mRNA), which are transcribed from protein-coding genes, meaning that the mRNAs can result in the production of particular proteins. A protein-coding gene consists of alternating sections called introns and exons. Exonic regions contain code for what a protein should be made up of, whereas the intronic regions can contain information on how to assemble the pieces.

% CVC: Work in progress
%A gene is a stretch of DNA that contains a genetic blueprint for a RNA molecule. The gene serves as a template for making an RNA molecule in a process called transcription. 
%The resulting RNA transcript can be protein-coding or non-coding. For this work we only consider the protein-coding genes, which consists of alternating sections called introns and exons. Exonic regions contain code for what a protein should be made up of, whereas the intronic regions can contain information on how to assemble the pieces.
%
%A protein-coding gene encodes protein-coding RNA molecules, the so-called messenger RNA (mRNA).
%
%This work only considers the protein-coding RNA molecules, the so-called messenger RNA (mRNA).
%% TODO: mention that there are about 20'000 protein-coding genes in the human body.
%
%The DNA that serves as a template for making a protein-coding RNA molecule is called a gene, and the complete set of an organism's genes is called its genome.
%
%The protein-coding part of the DNA is called a gene.

\textbf{RNA splicing} is a process that occurs in almost all organisms and results in the removal of intronic regions in the mRNA molecules, resulting in a mature mRNA. 
Splicing allows to create multiple variants of the same product, which can affect the enzymatic properties or localisation of the resulting product \cite{kelemen_functionalternativesplicing_2013}.

During \textbf{translation}, a chain of amino acids is synthesised from a mature mRNA transcript. 
% HT: in ribosomes. These units can be seen as small factories in which amino acids are assembled following the pattern contained in an mRNA molecule.
% RC: I can avoid talking about ribosomes
Every three nucleobases are translated into one of 21 different amino acids. The resulting chain of amino acids is folded up into a protein, the structure of which is determined by the sequence of different amino acids in the chain. In turn, its structure determines the functionality of the protein, which includes catalysing biochemical reactions, providing structure, and transportation of molecules.

\subsection{Cell types}
The functionality provided by a cell is defined (mostly) by the proteins of which it consists. One common approach to trying to understand the functionality of a cell is to observe which molecules are present in the cell and to associate those molecules with functionality.

\textit{Homo sapiens} like to categorise everything they encounter and so too have they conceptualised groups of cells called "cell types" according to their functionality. The concept of cell types eases reasoning about all aspects of biology, for instance, which cell types turn into (differentiate) or communicate with which other cell types, or how a cell type responds to a specific stimulation. Cells can be highly specialised toward performing a particular function (e.g. memory B cells accelerate immune response by remembering previously encountered pathogens), or they can maintain a strong ability to differentiate into other cell types.

\mycomment{do not only discuss differentiation, but instead any kind of developmental process; and mention 'developmental trajectory'.}
Cell differentiation is not an instant process; it is a continuous process in which a cell gradually produces the biochemical machinery required in order to fulfil a particular task. In this regard, it makes sense not only to reason about cell types but also about the transition states between cell types and the dynamic processes involved therein.

\mycomment{are dynamic processes defined? a dynamic process is just a biochemical reaction}

\subsection{Cell dynamics and gene regulation}
If cells are dynamic entities and can gradually produce the molecules needed to acquire new functionality, what is the process by which this happens? The mechanism by which this happens is called gene regulation. Some proteins (or other molecules such as micro RNAs) are capable of determining the rate at which a gene is transcribed (transcription rate). Such proteins are called transcription factors (TFs), and the genes they regulate are called their targets. Typically, one TF will regulate the transcription rate of many targets.

\mycomment{TODO: download database and calculate some statistics?}

@@

\mycomment{TODO: explain regulation mechanisms such as transcription factor binding sites?}
\subsection{Profiling single cells}
\mycomment{This sections goes greatly into detail about IHC and cytometry. Can the merging of the 'single-cell' and 'omics' be explained without going so much into detail?}
In order to understand a biological process, it is often quite helpful to be able to profile (i.e. observe) the biomolecular components involved therein. The single-cell "omics" technologies which we have at our fingertips today originated from the convergence of two different fields, \emph{"single-cell"} and \emph{"omics"}.

The earliest approaches for measuring the abundance of particular molecules in \emph{single cells} used the preferred instrument of every stereotypical biologist: the microscope.
Since it was developed by Coons et al. in 1941, immunohistochemistry (IHC) has been instrumental in visualising antigen-antibody proteins \cite{coons_immunologicalpropertiesantibody_1941}. In many multicellular organisms, antibodies and antigens serve as crucial communication tools as part of the organism's immune system. A cell can present a particular type of antigen on its cell surface, which allows a particular type of antibody to bind to it.

\mycomment{Rephrase the microscope sentence}

@@

Flow cytometry \cite{fulwyler_electronicseparationbiological_1965} is a technique which circumvents imaging and segmentation issues by having a steady stream of cells run through a laser and measuring the amount of light scattered from those cells. Flow cytometry technology enables to measure protein expression levels for millions of cells and tens of different antibodies.

@@

On the other side of the spectrum are the so-called "omics" technologies. "Omics"\footnote{The etymology of "omics" is quite interesting \cite{yadav_wholenesssuffixomics_2007}.} is a collective term for profiling all molecules of a particular type in a high-throughput manner. There are many types of "omics", but the most commonly used are the following. In genomics, all of an organism's genes are studied -- its whole genome. Transcriptomics and proteomics study the organisms RNA transcripts and proteins, respectively. A notable downside of traditional omics technologies is that in order to capture enough material an ensemble of cells needs to be profiled, and thus only the average expression levels are returned; thereby granting the technology the name "bulk" omics. If a subset of these cells contains unique patterns in expression levels, this pattern will be masked in the bulk population and is thus undetectable. Specific examples of omics technologies are next-generation sequencing, which can be used to determine the DNA sequence of an organism, and RNA sequencing, which profiles the sequences of RNA transcripts. By mapping the sequences of RNA transcripts to genes in the organisms DNA, a gene expression profile can be obtained.
\mycomment{Demonstrate the masking effect of bulk analyses.}

Transformative technological advances in microvolume sequencing allowed Tang et al. to analyse the transcriptome at single-cell resolution \cite{tang_mrnaseqwholetranscriptomeanalysis_2009}, thereby bringing single-cell biology and omics together to create single-cell omics (Figure \ref{fig:profiling_convergence}A). During the decade that followed, the number of single-cell omics technologies has skyrocketed, allowing to profile tens of thousands of cells (Figure \ref{fig:profiling_convergence}B) and measuring other levels of information such as proteomic expression levels (Figure \ref{fig:profiling_convergence}C).

\begin{figure}[htb!]
  \centering
  A. \includegraphics[width=.45\linewidth]{fig/profiling_convergence} 
  % also figure with cell numbers
  C. \includegraphics[width=.45\linewidth]{fig/multimodal} 
  % TODO: change image
  \caption{
    A. \textbf{Convergence of "Omics" Biology and Single-Cell Biology.} Technology that allows researchers to obtain genome-wide information from single cells is extending the boundaries of a field that has thus far been limited to the analyses of a select gene in eukaryotes. Image from Junker and van Oudenaarden (2014) \cite{junker_everycellspecial_2014}. 
    B. 
    C. scmultiomics \cite{moudgil_multimodalscrnaseq_2019}. 
    % TODO: rewrite the description
  }
  \label{fig:profiling_convergence}
\end{figure}

@@

\begin{figure}[htb!]
  \centering\includegraphics[width=.5\linewidth]{fig/singlecell_tools}
  \centering\includegraphics[width=.3\linewidth]{fig/scrnatools}
  % TODO: change image
  \caption{\textbf{A. Single-cell omics allows for many new types of computational approaches.} Figure adapted from Wagner et al. (2016) \cite{wagner_revealingvectorscellular_2016}.
    B. Zappia et al. (2018) \cite{zappia_exploringsinglecellrnaseq_2018}
    % TODO: rewrite the description
  }
  \label{fig:comp_tools}
\end{figure}

@@

There are many ways of classifying DR methods \cite{engel_surveydimensionreduction_2012}, but this work will use the following main categories: feature projection-based and manifold learning. 
Projection-based DR methods aim to perform a linear transformation of the data while preserving the pairwise distances between samples as much as possible. Examples of commonly used projection-based DR methods in single-cell omics are PCA and MDS. 
Manifold learning methods are methods which reconstruct a higher-order structure in the original space (e.g. a graph or a grid), visualising the structure in a lower-dimensional space, and mapping the original samples to the lower-dimensional space. Manifold learning can be an iterative optimisation process using a predefined criterion. Examples of manifold learning techniques are t-SNE, Diffusion Maps and UMAP.

@@

\subsection{Trajectory inference}
Single-cell omics data provide new opportunities for studying cellular dynamic processes, such as the cell cycle, cell differentiation and cell activation \cite{tanay_scalingsinglecellgenomics_2017,etzrodt_quantitativesinglecellapproaches_2014}. 
Trajectory inference (TI) is a new category of computational tools used to offer an unbiased and transcriptome-wide understanding of a dynamic process \cite{tanay_scalingsinglecellgenomics_2017,cannoodt_computationalmethodstrajectory_2016}.

Technological advancements in single-cell omics allow studying a dynamic process in a high-throughput manner. 
This raises concerns regarding biological fundamentals, such as how to define cell types or transitions between them \cite{etzrodt_quantitativesinglecellapproaches_2014,tanay_scalingsinglecellgenomics_2017}. 
Trajectory inference (TI) methods aim to give insight into a dynamic process by inferring a trajectory from omics profiles of cells in which the dynamic process takes place \cite{cannoodt_computationalmethodstrajectory_2016}. 
TI has two objectives: to reconstruct the topology of the dynamic process (e.g. is it linear, cyclical, bifurcating), and to determine the position of each cell along the topology. Some TI methods assumes that the user knows the topology beforehand and only focuses on ordering the cells along a predefined topology.

The dataset can be a single snapshot of a mixture of cells in different stages, or a set of samples collected at different time points (Figure \ref{fig:trajectory_inference}A). Typically, TI methods first analyse similarities between cells, optionally infer the topology of the underlying process, and finally order cells along that trajectory (Figure \ref{fig:trajectory_inference}B). The second step can be optional, as some methods assume a specific topology beforehand.
TI methods allow the identification of new subsets of cells, delineation of a differentiation tree, and characterisation of the main driver genes along a state transition (Figure \ref{fig:trajectory_inference}C). Current applications of TI focus on specific subsets of cells, but ongoing efforts to construct transcriptomic catalogs of whole organisms \cite{regev_humancellatlas_2017,han_mappingmousecell_2018,schaum_singlecelltranscriptomics20_2018} underline the urgency for accurate, scalable \cite{aibar_scenicsinglecellregulatory_2017,angerer_singlecellsmake_2017} and user-friendly TI methods.

@@

\subsection{Gene regulatory network inference}
Gene regulatory network inference, or network inference (NI) for short, is a type of computational analysis where thousands of transcriptomic profiles are analysed together in order to infer the regulatory interactions between transcription factors and genes. This topic already received much attention with the advent of bulk omics (before single-cell omics). These efforts culminated in several DREAM competitions assessing the performance of 29 different NI methods \cite{marbach_revealingstrengthsweaknesses_2010, marbach_wisdomcrowdsrobust_2012}.

@@

The advent of single-cell omics has made scientists wonder whether now is the time to revisit network inference \cite{stegle_computationalanalyticalchallenges_2015}. 
One of the main advantages of single-cell omics is the ability to quantify the exact cellular state of thousands of cells per experiment. The heterogeneity between cells caused by naturally occurring biological randomness \cite{padovan-merhar_usingvariabilitygene_2013} can be exploited to infer regulatory interactions between TFs and their target genes at much lower costs (see Figure \ref{fig:bulk_vs_singlecell}).
In this setting, heterogeneity in the cell population eases network inference, rather than mask condition-specific expression patterns in regulatory interactions.

\section{Research context and objectives}
Recent technological advancements in profiling single cells are having significant repercussions in many fields of biology. Profiling thousands of individual cells in a genome-wide manner provides opportunities to study cell heterogeneity and dynamics, for example inferring mechanisms for cellular development or intercellular communication. 
Hundreds of new software tools were developed \cite{zappia_exploringsinglecellrnaseq_2018} to perform these new types of analyses, or to fit existing analytical tools to deal with new data characteristics (e.g. differential expression, dimensionality reduction, normalisation).

>>>>

\section{The cell}
The cell is the smallest unit of life, of which all known living organisms are composed. Every cell houses a plethora of biomolecular processes that allows it to continuously adapt to changes in its environment. Due to the dynamic nature of these processes, it can be very challenging to comprehend the cellular response to a signal. A reductionist approach to understanding a complex biological system is to study the biochemical components of which it is comprised\cite{brigandt_reductionismbiology_2017}.

Recent advances in experimental technologies are playing a crucial role in reductionist biology, allowing to measure the abundance of thousands of different biochemical molecules in tens of thousands of individual cells. With it comes the challenge of analysing large amounts of data that are not easily interpretable by hand. The sheer volume of the data generated from such highly-integrative and high-throughput experiments are not the only reason why they are so challenging to interpret. For instance, the generated data contains high levels of noise arising from inherent biomolecular stochasticity in the cells and from the experimental profiling techniques used, as well as batch effects arising from differences between donors and labs\cite{hon_humancellatlas_2018}. Biologists thus turn to computer scientists
to develop new tools to tackle these problems and help them to extract meaningful biological insights from the data. In this work, incremental contributions were made to the field in order to be able to address the aforementioned problems in a more comprehensive context.

Observing the biomolecular insides of cells can ultimately provide fundamental understanding into the processes that govern these cells and help uncover novel approaches for disease diagnosis, prognosis, and treatment. For example, the Human Cell Atlas (HCA) consortium\cite{regev_humancellatlas_2018} has set out to develop a comprehensive reference map of all the different types of cells in the human body. Experts in the field often metaphorically describe the HCA initiative as aiming to develop a 'Google Maps' of the human body. Even in its infancy, the HCA has profiled 3.8 million cells from 248 donors across 42 labs\cite{humancellatlasconsortium_humancellatlas_2018}, and this number is likely to increase well above one hundred million.

@@

\subsection{The origin of life and the RNA world}
The discovery of the double helix shape of deoxyribonucleic acid (DNA)\cite{watson_molecularstructurenucleic_1953} is often considered the pivot point in our understanding of the origin of life and evolution. By now, it is well known that DNA serves as a medium for storing the genetic information required to reproduce a whole organism. With other words, the DNA of an organism contains the complete set of instructions required to build all of the biomolecular machinery present in its body.

Life (or cells) did not originate from DNA, however. A widely-accepted hypothesis states that life originates from its lesser-known cousin, ribonucleic acid (RNA). According to the RNA world hypothesis\cite{alberts_rnaworldorigins_2002}, the very first primitive cells used RNA both to store genetic information and to perform the chemical reactions required to sustain themselves (Figure \ref{fig:rnaworld}). Only later did cells develop the ability to use the more chemically stable DNA molecules to self-sustain in a process commonly referred to as the central dogma.

\begin{figure}[htb!]
	\centering\includegraphics[width=.5\linewidth]{fig/rnaworld} % TODO: change image
	\caption{
		\textbf{RNA world.} The postulated rise and fall of the RNA world during the evolution of life, from early self-replicating RNA to complex, RNA-controlled metabolism, to the invention of translation, followed by diversification of all modern branches of life. Image from Horning (2011)\cite{horning_rnaworld_2011}. % TODO: expand the description
	}
	\label{fig:rnaworld}
\end{figure}

The central dogma describes the general flow of genetic information in almost all existing living cells: DNA is decoded to RNA, which in turn encodes proteins. Main processes involved in the central dogma are \textbf{transcription}, \textbf{splicing}, and \textbf{translation} (Figure \ref{fig:centraldogma}).

@@

During the process of \textbf{transcription} that takes place in the cell nucleus, a complementary RNA copy is transcribed from the template DNA. The initial RNA transcript is a precursor messenger RNA (pre-mRNA) that needs to undergo series of maturation steps to ultimately form the mature messenger RNA (mRNA). This maturation includes pre-mRNA \textbf{splicing} to remove non-protein coding intervening sequences (the introns) and to join the neighbouring protein-coding sequences (the exons). A single pre-mRNA can be alternatively spliced to generate multiple forms of mRNAs that will result in the production of multiple protein isoforms. This process of alternative splicing is essential to generate over a 100.000 different proteins starting from a rather small number of genes (20.000-25.000).

 The mature mRNA is then transported to the cytoplasm, where it engages with ribosomes to initiate \textbf{translation}. During this highly evolutionary conserved process, a chain of amino acids, known as the protein building blocks, is being synthesised. Each amino acid is specified by three nucleotides (a codon) in the mRNA, according to a nearly universal genetic code. After being released by the ribosomes, the translation product undergoes a variety of chemical modifications to form the final folded protein, the structure of which is determined by the sequence of different amino acids in the chain. In addition, polypeptides may be cleaved to yield more than one active polypeptide product. The structure of a protein determines its functionality, which includes catalysing biochemical reactions, providing structure, and transportation of molecules.
 % Ref: Matera AG, Wang Z. A day in the life of the spliceosome. Nature reviews Molecular cell biology. 2014 Feb;15(2):108. 
 % Ref: Book: Human Molecular Genetics, Strachan and Read

\subsection{Cell types}
\textit{Homo sapiens} like to categorise everything they encounter, and so too have they conceptualised groups of cells called "cell types". The human body contains more than 200 different cell types that are classified into four groups: epithelial, connective, muscle, and nervous. This however, is a major underestimation of the real number of cell types. Neurons, for instance, that are known to be extremely diverse, are estimated to reach numbers above 10,000 different types.
% Ref: Strachan and Read

The concept of cell types eases reasoning and our understanding about many aspects of biology (e.g. the process of cell differentiation, cell-cell communication, cellular response to certain stimuli). Some cells are known to be highly specialised toward performing a particular function (e.g. memory B cells accelerate immune response by remembering previously encountered pathogens), or they can maintain a strong ability to differentiate into other cell types.

One common approach for understanding the functionality of a particular cell is to observe which molecules are present in the cell and to associate those set of molecules with functionality. Taking a snap shot of the protein or RNA transcript content in a particular cell, might already provide us with major insights into its functionality. However, in order to fulfil a particular task, the biochemical machinery of the cell gradually changes over time. Therefore it is highly informative to also consider the transition states between cell types and the dynamic processes involved therein.

\subsection{Cell dynamics and gene regulation}
Cells are dynamic entities that can gradually produce the molecules needed to acquire new functionality. The variation within and between cells happens because of differences in gene expression. The mechanism by which this happens is called gene regulation, and takes place primarily at the level of transcription. According to the needs of a cell, different genes are being transcribed. Housekeeping genes are being expressed in essentially every cell, while other genes are cell type or tissue specific and may be expressed at certain developmental stages.
%ref: Strachan and Read
% CVC

Some proteins (or other molecules such as micro RNAs) are capable of determining the rate at which a gene is transcribed (transcription rate). Such proteins are called transcription factors (TFs), and the genes they regulate are called their targets. Typically, one TF will regulate the transcription rate of many targets.

@@

\subsection{Profiling single cells}
In order to understand a biological process, it is often quite helpful to be able to profile (i.e. observe) the biomolecular components involved therein. The single-cell "omics" technologies which we have at our fingertips today originated from the convergence of two different fields, \emph{"single-cell"} and \emph{"omics"}.

\subsubsection{Single-cell}
The earliest approaches for measuring the abundance of particular molecules in \emph{single cells} used the preferred instrument of every stereotypical biologist: the microscope.
Since it was developed by Coons et al. in 1941, immunohistochemistry (IHC) has been instrumental in visualising antigen-antibody proteins\cite{coons_immunologicalpropertiesantibody_1941}. In many multicellular organisms, antibodies and antigens serve as crucial communication tools as part of the organism's immune system. A cell can present a particular type of antigen on its cell surface, which allows a particular type of antibody to bind to it.

\mycomment{Rephrase the microscope sentence (dino joke?)}

@@

Flow cytometry\cite{fulwyler_electronicseparationbiological_1965} is a technique which circumvents imaging and segmentation issues by having a steady stream of cells run through a laser and measuring the amount of light scattered from those cells. Flow cytometry technology enables to measure protein expression levels for millions of cells and tens of different antibodies.

@@

\subsubsection{Omics}
On the other side of the spectrum are the so-called "omics" technologies. "Omics"\footnote{The etymology of "omics" is quite interesting\cite{yadav_wholenesssuffixomics_2007}.} is a collective term for profiling all molecules of a particular type in a high-throughput manner. There are many types of "omics", but the most commonly used are the following. In genomics, all of an organism's genes are studied -- its whole genome. Transcriptomics and proteomics study the organisms RNA transcripts and proteins, respectively. A notable downside of traditional omics technologies is that in order to capture enough material an ensemble of cells needs to be profiled, and thus only the average expression levels are returned; thereby granting the technology the name "bulk" omics. If a subset of these cells contains unique patterns in expression levels, this pattern will be masked in the bulk population and is thus undetectable. Specific examples of omics technologies are next-generation sequencing, which can be used to determine the DNA sequence of an organism, and RNA sequencing, which profiles the sequences of RNA transcripts. By mapping the sequences of RNA transcripts to genes in the organisms DNA, a gene expression profile can be obtained.
\mycomment{Demonstrate the masking effect of bulk analyses.}

\subsubsection{Single-cell omics}
Transformative technological advances in microvolume sequencing allowed Tang et al. to analyse the transcriptome at single-cell resolution\cite{tang_mrnaseqwholetranscriptomeanalysis_2009}, thereby bringing single-cell biology and omics together to create single-cell omics (Figure \ref{fig:profiling_convergence}A). During the decade that followed, the number of single-cell omics technologies has skyrocketed, allowing to profile tens of thousands of cells (Figure \ref{fig:profiling_convergence}B) and measuring other levels of information such as proteomic expression levels (Figure \ref{fig:profiling_convergence}C).

\begin{figure}[htb!]
  \centering
  A. \includegraphics[width=.45\linewidth]{fig/profiling_convergence} 
  % also figure with cell numbers
  C. \includegraphics[width=.45\linewidth]{fig/multimodal} 
  % TODO: change image
  \caption{
    A. \textbf{Convergence of "Omics" Biology and Single-Cell Biology.} Technology that allows researchers to obtain genome-wide information from single cells is extending the boundaries of a field that has thus far been limited to the analyses of a select gene in eukaryotes. Image from Junker and van Oudenaarden (2014)\cite{junker_everycellspecial_2014}. 
    B. 
    C. scmultiomics\cite{moudgil_multimodalscrnaseq_2019}. 
    % TODO: rewrite the description
  }
  \label{fig:profiling_convergence}
\end{figure}

@@

\begin{figure}[htb!]
  \centering\includegraphics[width=.5\linewidth]{fig/singlecell_tools}
  \centering\includegraphics[width=.3\linewidth]{fig/scrnatools}
  % TODO: change image
  \caption{\textbf{A. Single-cell omics allows for many new types of computational approaches.} Figure adapted from Wagner et al. (2016)\cite{wagner_revealingvectorscellular_2016}.
    B. Zappia et al. (2018)\cite{zappia_exploringsinglecellrnaseq_2018}
    % TODO: rewrite the description
  }
  \label{fig:comp_tools}
\end{figure}

@@

There are many ways of classifying DR methods\cite{engel_surveydimensionreduction_2012}, but this work will use the following main categories: feature projection-based and manifold learning. 
Projection-based DR methods aim to perform a linear transformation of the data while preserving the pairwise distances between samples as much as possible. Examples of commonly used projection-based DR methods in single-cell omics are PCA and MDS. 
Manifold learning methods are methods which reconstruct a higher-order structure in the original space (e.g. a graph or a grid), visualising the structure in a lower-dimensional space, and mapping the original samples to the lower-dimensional space. Manifold learning can be an iterative optimisation process using a predefined criterion. Examples of manifold learning techniques are t-SNE, Diffusion Maps and UMAP.

@@

\subsection{Trajectory inference}
Single-cell omics data provide new opportunities for studying cellular dynamic processes, such as the cell cycle, cell differentiation and cell activation\cite{tanay_scalingsinglecellgenomics_2017,etzrodt_quantitativesinglecellapproaches_2014}. 
Trajectory inference (TI) is a new category of computational tools used to offer an unbiased and transcriptome-wide understanding of a dynamic process\cite{tanay_scalingsinglecellgenomics_2017,cannoodt_computationalmethodstrajectory_2016}.

Technological advancements in single-cell omics allow studying a dynamic process in a high-throughput manner. 
This raises concerns regarding biological fundamentals, such as how to define cell types or transitions between them\cite{etzrodt_quantitativesinglecellapproaches_2014,tanay_scalingsinglecellgenomics_2017}. 
Trajectory inference (TI) methods aim to give insight into a dynamic process by inferring a trajectory from omics profiles of cells in which the dynamic process takes place\cite{cannoodt_computationalmethodstrajectory_2016}. 
TI has two objectives: to reconstruct the topology of the dynamic process (e.g. is it linear, cyclical, bifurcating), and to determine the position of each cell along the topology. Some TI methods assumes that the user knows the topology beforehand and only focuses on ordering the cells along a predefined topology.

The dataset can be a single snapshot of a mixture of cells in different stages, or a set of samples collected at different time points (Figure \ref{fig:trajectory_inference}A). Typically, TI methods first analyse similarities between cells, optionally infer the topology of the underlying process, and finally order cells along that trajectory (Figure \ref{fig:trajectory_inference}B). The second step can be optional, as some methods assume a specific topology beforehand.
TI methods allow the identification of new subsets of cells, delineation of a differentiation tree, and characterisation of the main driver genes along a state transition (Figure \ref{fig:trajectory_inference}C). Current applications of TI focus on specific subsets of cells, but ongoing efforts to construct transcriptomic catalogs of whole organisms\cite{regev_humancellatlas_2017,han_mappingmousecell_2018,schaum_singlecelltranscriptomics20_2018} underline the urgency for accurate, scalable\cite{aibar_scenicsinglecellregulatory_2017,angerer_singlecellsmake_2017} and user-friendly TI methods.

@@

\subsection{Gene regulatory network inference}
Gene regulatory network inference, or network inference (NI) for short, is a type of computational analysis where thousands of transcriptomic profiles are analysed together in order to infer the regulatory interactions between transcription factors and genes. This topic already received much attention with the advent of bulk omics (before single-cell omics). These efforts culminated in several DREAM competitions assessing the performance of 29 different NI methods\cite{marbach_revealingstrengthsweaknesses_2010, marbach_wisdomcrowdsrobust_2012}.

@@

The advent of single-cell omics has made scientists wonder whether now is the time to revisit network inference\cite{stegle_computationalanalyticalchallenges_2015}. 
One of the main advantages of single-cell omics is the ability to quantify the exact cellular state of thousands of cells per experiment. The heterogeneity between cells caused by naturally occurring biological randomness\cite{padovan-merhar_usingvariabilitygene_2013} can be exploited to infer regulatory interactions between TFs and their target genes at much lower costs (see Figure \ref{fig:bulk_vs_singlecell}).
In this setting, heterogeneity in the cell population eases network inference, rather than mask condition-specific expression patterns in regulatory interactions.

\section{Research context and objectives}
Recent technological advancements in profiling single cells are having significant repercussions in many fields of biology. Profiling thousands of individual cells in a genome-wide manner provides opportunities to study cell heterogeneity and dynamics, for example inferring mechanisms for cellular development or intercellular communication. 
Hundreds of new software tools were developed\cite{zappia_exploringsinglecellrnaseq_2018} to perform these new types of analyses, or to fit existing analytical tools to deal with new data characteristics (e.g. differential expression, dimensionality reduction, normalisation).
